---
title: "R_BFS_hypervolume_final"
author: "A Taylor, B Nguyen"
date: "2023-06-10"
output:
  html_document: default
---

```{r setup, include=FALSE}
setwd("/Users/andrewtaylor/Desktop/R_Projects/R_BFS_hypervolume/")
knitr::opts_chunk$set(echo = TRUE)
```

## Importing & preparing data

We will import the occurrence data used for the Bluntface Shiner Maxent models, which will become the basis for quantifying niche hypervolume and overlap statistics.

Data will be arranged into two dataframes, one for western populations ("ARW") and one for eastern populations ("LMT"). Columns will consist of the environmental covariates found to be most influential across Maxent models.

##Standardizing data ranges for final comparisons

Hypervolume methodology is sometimes biased when some variables have much larger range/SD than others (e.g., TotDA in our dataset). One way to address this is to standardize each variable by transforming it with z-scores. We will start with the entire BFS range's data together, standardize each variable's range with z-scores, then split into ARW and LMT for comparisons.

```{r data import & standardizing, message=TRUE, warning=TRUE, paged.print=TRUE}
#ALL is the combined dataset of ARW and LMT
ALL <- read.csv("BFSAllPresOnly.csv", header = T, na.strings= "-9999")
head(ALL)
#Remove duplicated stream reach occurrences by keeping one of each COMID number.
ALL <- ALL[!duplicated(ALL$COMID),]
#Retain only columns of influential covariates
#plus region, to sort later
ALL <- ALL[, c(2,7,8,10,14)]
#Remove any records w/ NA 
ALL <-na.omit(ALL)
head(ALL)
dim(ALL)

#Z-score scale columns with covariates
ALL['ElevCat'] <- scale(ALL['ElevCat'])
ALL['ClayCat'] <- scale(ALL['ClayCat'])
ALL['RckDepCat'] <- scale(ALL['RckDepCat'])
ALL['TotDASqKM'] <- scale(ALL['TotDASqKM'])

#Seperate into ARW and LMT for scaled analysis
ARWz = ALL[ALL$Region=="West",c(2,3,4,5)]
LMTz = ALL[ALL$Region=="East",c(2,3,4,5)]
```

## Building Single, Standardized Range Hypervolumes

```{r building standardized hypervolumes , echo=FALSE}
require(hypervolume)
ARWz_hv = hypervolume(ARWz, name = 'ARW')
summary(ARWz_hv)
plot(ARWz_hv)

LMTz_hv = hypervolume(LMTz, name = 'LMT')
summary(LMTz_hv)
plot(LMTz_hv)

get_volume(ARWz_hv)
get_volume(LMTz_hv)
```

## Building Standardized Hypervolume Set and Comparisons

```{r hypervolume comparisons standardized, echo=FALSE, message=FALSE}
#Compare set of two species
hvz_set <- hypervolume_set(ARWz_hv, LMTz_hv, num.points.max = NULL, 
                          verbose = TRUE, check.memory = FALSE, distance.factor = 1)
#Raw data overlap stats
hypervolume_overlap_statistics(hvz_set)

#compare two hypervolumes in a plot
plot(hvz_set)

###Resample each hypervolume to create a mean and variation around it, so 
###they can be more accurately compared w/ statistics
```

```{r resamples for CI estimation, echo=FALSE, message=FALSE, results=FALSE}
#generates confidenceintervals from quantilesof 20*20 overlaps 
#hiding this from markdown because it takes forever and very verbose!
path1z=hypervolume_resample("ARWz_hv_meanHV",ARWz_hv,n=20, method = "bootstrap") 
path2z=hypervolume_resample("LMTz_hv_meanHV",LMTz_hv,n=20, method = "bootstrap") 
```

###Confidence Intervals for Overlap Metrics

```{r CI estimates, echo=FALSE, message=FALSE}
resultz=hypervolume_overlap_confidence(path1z,path2z) 
#confidence intervals for similarity scores
print(resultz["jaccard"])
print(resultz["sorensen"])
#fraction of ARW hypervolume that is unique (first hypervol)
print(resultz["frac_unique_1"])
#fraction of LMT hypervolume that is unique (second hypervol)
print(resultz["frac_unique_2"])
#matrix of overlap stats used
print(resultz["distribution"])
```

```{r finished notification, echo=FALSE, message=FALSE, results=FALSE}
#Holla at ya boi when it is done
if(!require(devtools)) {install.packages("devtools")}
devtools::install_github("brooke-watson/BRRR")
library(BRRR)
#JUNIOR MAFIA
skrrrahh("biggie")
```

#Customizing hypervolume plot colors and legend
```{r customized hypervolume plot and legend, echo=FALSE, message=FALSE}

#colorbrewer2 suggested pallette
newcolors2 <- c("#fc8d59", "#91bfdb","#fee090", "#e0f3f8", "#d73027", "#4575b4")

#Make new plot
hyperplot2 <- plot(hvz_set, show.3d=FALSE, plot.3d.axes.id=NULL,
               show.axes=TRUE, show.frame=TRUE,
               show.random=TRUE, show.density=TRUE,show.data=TRUE,
               names=NULL, show.legend=FALSE, limits=NULL, 
               show.contour=TRUE, contour.lwd=1.5, 
               contour.type='kde', 
               contour.alphahull.alpha=0.25,
               contour.ball.radius.factor=1, 
               contour.kde.level=1e-04,
               contour.raster.resolution=20,
               show.centroid=TRUE, cex.centroid=1.5,
               colors=newcolors2, 
               point.alpha.min=0.2, point.dark.factor=0.5,
               cex.random=0.5,cex.data=0.75,cex.axis=0.75,cex.names=1.0,cex.legend=1.0,
               num.points.max.data = 1000, num.points.max.random = 2000, reshuffle=TRUE,
               plot.function.additional=NULL,
               verbose=FALSE)

#Add custom legend
legend(x=0,y=0.2,  # Coordinates (x also accepts keywords)
       legend = c("LMT", "ARW", "Intersection", "Union", "Unique ARW", "Unique LMT"), # Vector with the name of each group
       #fill,   # Creates boxes in the legend with the specified colors
       col = newcolors2, # Color of lines or symbols
       box.lty=1, box.lwd=2, box.col="black", #Lengend box border
       #border = "black", # Fill box border color
       lty=1, lwd=1.5,         # Line type and width
       pch=16,              # Add pch symbols to legend lines or boxes
       #bty = "n",        # Box type (bty = "n" removes the box)
       bg = "white",    # Background color of the legend
       #box.lwd = par("lwd"), # Legend box line width
       #box.lty = par("lty"), # Legend box line type
       #box.col = par("fg"),  # Legend box line color
       cex = 0.5,          # Legend size
       horiz = FALSE,     # Horizontal (TRUE) or vertical (FALSE) legend
       title = "Legend",
       text.width = 0.1, text.font=4
)      

```